---
title: "Setup and data (Chapter 4)"
---

::: {.callout-tip title="How to use this document"}
This is the code from the handbook chapter with the explanatory text removed. Run the chunks in order from the top; later chunks depend on objects created earlier. The numbered notes under a chunk explain the marked lines. The full discussion of each stage is in the handbook chapter.
:::

::: {.callout-note title="Main takeaways"}
- All worked examples in Chapters 5–7 use a single dataset, the Trends in International Mathematics and Science Study (TIMSS) 2015 Canada Grade 4 (`timss`), loaded from `timss_df.rds`.
- The data has a multilevel structure (students nested within teachers nested within schools).
- Three "treatment" indicators in the dataset map to the three multilevel designs covered in this handbook: `multisite_treatment` (MIAD), `cluster_treatment` (CAD), and `multisite_cluster_treatment` (MCAD).
- The treatment indicators were created from observed covariates and do not represent real interventions.
:::

## R setup and packages

### Installing packages

```{r}
#| eval: false

install.packages(c(
  "tidyverse",
  "MatchIt",
  "matchMulti",
  "CMatching",
  "optmatch",
  "gbm",
  "dbarts",
  "cobalt",
  "lme4",
  "performance",
  "marginaleffects",
  "sensemakr",
  "gt"
))
```

### Data manipulation and visualization

#### `tidyverse`

```{r}
#| message: false
#| warning: false

library(tidyverse)

#   )
```

### Propensity score matching

#### `MatchIt`

```{r}
#| message: false
#| warning: false

library(MatchIt)

# matched_df <- match.data(match_out)
```

#### `matchMulti`

```{r}
#| message: false
#| warning: false

library(matchMulti)

# )
```

#### `CMatching`

```{r}
#| message: false
#| warning: false

library(CMatching)

# )
```

#### `gbm`

```{r}
#| message: false
#| warning: false

library(gbm)

# )
```

#### `dbarts`

```{r}
#| message: false
#| warning: false

library(dbarts)

# )
```

### Balance assessment

#### `cobalt`

```{r}
#| message: false
#| warning: false

library(cobalt)

#           thresholds = c(m = .25, v = 2))
```

### Multilevel modeling

#### `lme4`

```{r}
#| message: false
#| warning: false

library(lme4)

# VarCorr(model)  # Extract variance components
```

#### `performance`

```{r}
#| message: false
#| warning: false

library(performance)

# icc(model)
```

### Treatment effect estimation

#### `marginaleffects`

```{r}
#| message: false
#| warning: false

library(marginaleffects)

# )
```

### Sensitivity analysis

#### `sensemakr`

```{r}
#| message: false
#| warning: false

library(sensemakr)

# plot(sens, sensitivity.of = "estimate")
```

### Table formatting

#### `gt`

```{r}
#| message: false
#| warning: false

library(gt)

#   fmt_number(columns = everything(), decimals = 1)
```

## Data file description

```{r}
#| message: false
#| warning: false

timss <- readRDS("data/timss_df.rds")
```

## Variable definitions

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-var-defs
#| tbl-cap: "Variable definitions grouped by category."

variable_defs <- tibble::tribble(
  ~category           , ~name                         , ~definition                                         , ~level    , ~coding                           , ~response_labels                                     ,
  # IDs
  "ID"                , "school_id"                   , "School/site identifier"                            , "School"  , "Integer ID"                      , ""                                                   ,
  "ID"                , "teacher_id"                  , "Teacher identifier"                                , "Teacher" , "Integer ID"                      , ""                                                   ,
  "ID"                , "class_id"                    , "Classroom identifier"                              , "Class"   , "Integer ID"                      , ""                                                   ,
  "ID"                , "student_id"                  , "Student identifier"                                , "Student" , "Integer ID"                      , ""                                                   ,
  # Outcome
  "Outcome"           , "math_score"                  , "TIMSS math achievement score"               , "Student" , "Continuous"                      , ""                                                   ,
  # Student covariates
  "Student covariate" , "student_sex"                 , "Student sex indicator"                             , "Student" , "0/1"                             , "1 = male; 0 = female"                               ,
  "Student covariate" , "student_age"                 , "Student age"                                       , "Student" , "Continuous"                      , ""                                                   ,
  "Student covariate" , "student_math_conf"           , "Math confidence scale"                             , "Student" , "Continuous"                      , ""                                                   ,
  "Student covariate" , "student_like_math"           , "Liking math scale"                                 , "Student" , "Continuous"                      , ""                                                   ,
  "Student covariate" , "attend_preschool"            , "Attended preschool"                                , "Student" , "0/1"                             , "1 = attended preschool; 0 = did not attend"                                                   ,
  "Student covariate" , "extra_lessons_freq"          , "Extra math lessons attendance (duration)"               , "Student" , "Ordered (4 levels)"              , "0 = did not attend; 3 = more than 8 months"         ,
  "Student covariate" , "dad_edu"                     , "Father's education"                                , "Student" , "Ordered (5 levels)"              , "1 = lower secondary or below; 5 = postgraduate"             ,
  "Student covariate" , "mom_edu"                     , "Mother's education"                                , "Student" , "Ordered (5 levels)"              , "1 = lower secondary or below; 5 = postgraduate"             ,
  # Teacher covariates
  "Teacher covariate" , "teacher_sex"                 , "Teacher sex indicator"                             , "Teacher" , "0/1"                             , "1 = male; 0 = female"                               ,
  "Teacher covariate" , "teacher_age"                 , "Teacher age"                                       , "Teacher" , "Continuous (category midpoints)" , "Age range midpoints in years"                       ,
  "Teacher covariate" , "teacher_years_taught"        , "Years of teaching experience"                      , "Teacher" , "Continuous"                      , "Whole years of teaching experience"                 ,
  "Teacher covariate" , "teacher_edu"                 , "Teacher education level"                           , "Teacher" , "Ordered (3 levels)"              , "1 = bachelor's; 2 = master's; 3 = doctoral (none in this sample)" ,
  "Teacher covariate" , "teacher_job_satisfaction"    , "Teacher job satisfaction"                          , "Teacher" , "Ordered (3 levels)"              , "1 = less than satisfied; 3 = very satisfied"        ,
  "Teacher covariate" , "teacher_math_major"          , "Teacher math major indicator"                      , "Teacher" , "0/1"                             , "1 = math major; 0 = not a math major"                                                   ,
  # School covariates
  "School covariate"  , "school_locale"               , "School locale category"                            , "School"  , "Ordered (5 levels)"              , "0 = rural; 4 = urban (no rural schools in this sample)" ,
  "School covariate"  , "school_pop"                  , "Population of the school's community/area"         , "School"  , "Continuous (category midpoints)" , "Mean of population range midpoints (number of people)"      ,
  "School covariate"  , "school_econ_pct"             , "Economically disadvantaged percentage"             , "School"  , "Continuous (category midpoints)" , "Mean of percentage range midpoints"                         ,
  "School covariate"  , "school_shortage"             , "Instruction affected by resource shortages scale"  , "School"  , "Continuous"                      , "Higher = instruction less affected by shortages"    ,
  "School covariate"  , "school_discipline"           , "School discipline/climate index"                   , "School"  , "Continuous"                      , ""                                                   ,
  "School covariate"  , "school_place_assist"         , "School provides students assistance with schoolwork" , "School"  , "0/1"                             , "1 = yes; 0 = no"                                                   ,
  "School covariate"  , "school_achieve_used_math"    , "School uses students' math achievement data" , "School"  , "0/1"                             , "1 = yes; 0 = no"                                                   ,
  # Treatment indicators
  "Treatment"         , "multisite_treatment"         , "Student receives extra lessons in math"           , "Student" , "0/1"                             , "1 = treated; 0 = comparison"                        ,
  "Treatment"         , "multisite_cluster_treatment" , "Student's teacher majored in math"          , "Teacher" , "0/1"                             , "1 = treated; 0 = comparison"                        ,
  "Treatment"         , "cluster_treatment"           , "School least affected by resource shortages (top third of schools)" , "School"  , "0/1"                             , "1 = treated; 0 = comparison"
)

variable_defs |>
  gt(groupname_col = "category") |>
  cols_label(
    name            = "Variable",
    definition      = "Definition",
    level           = "Level",
    coding          = "Coding",
    response_labels = "Response labels"
  ) |>
  tab_header(
    title    = "Variable definitions",
    subtitle = "Data: TIMSS 2015 Grade 4 (Canada)"
  ) |>
  tab_style(
    style     = cell_text(font = "monospace"),
    locations = cells_body(columns = name)
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_title()
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) |>
  tab_style(
    style     = cell_text(color = "gray70"),
    locations = cells_body(columns = response_labels, rows = response_labels == "")
  )
```

## Treatment indicators

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-trt-defs
#| tbl-cap: "Treatment indicator definitions."

treatment_defs <- tibble::tribble(
  ~name                         , ~definition                                         , ~level    , ~coding ,
  "multisite_treatment"         , "Student receives extra lessons in math"           , "Student" , "0/1"   ,
  "multisite_cluster_treatment" , "Student's teacher majored in math"          , "Teacher" , "0/1"   ,
  "cluster_treatment"           , "School least affected by resource shortages (top third of schools)" , "School"  , "0/1"
)

treatment_defs |>
  gt() |>
  cols_label(
    name = "Variable",
    definition = "Definition",
    level = "Level",
    coding = "Coding"
  ) |>
  tab_header(
    title = "Treatment indicators",
    subtitle = "Data: TIMSS 2015 Grade 4 (Canada)"
  ) |>
  tab_style(
    style = cell_text(font = "monospace"),
    locations = cells_body(columns = name)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_title()
  )
```

## Data structure and multilevel nesting

```{r}
#| message: false
#| warning: false
#| label: tbl-unit-counts
#| tbl-cap: "Unit counts by level."

n_students <- nrow(timss)
n_teachers <- n_distinct(timss$teacher_id)
n_schools  <- n_distinct(timss$school_id)

tibble(
  Level = c("Students", "Teachers", "Schools"),
  N     = c(n_students, n_teachers, n_schools)
) |>
  gt() |>
  tab_header(
    title    = "Multilevel structure of the analytic dataset",
    subtitle = "Data: TIMSS 2015 Grade 4 (Canada)"
  ) |>
  fmt_number(columns = N, decimals = 0) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_title()
  )
```

```{r}
#| message: false
#| warning: false
#| label: tbl-cluster-sizes
#| tbl-cap: "Cluster size variation."

students_per_teacher <- timss |>
  count(teacher_id, name = "n") |>
  summarise(Mean = mean(n), SD = sd(n), Min = min(n), Max = max(n)) |>
  mutate(Relationship = "Students per teacher")

teachers_per_school <- timss |>
  distinct(school_id, teacher_id) |>
  count(school_id, name = "n") |>
  summarise(Mean = mean(n), SD = sd(n), Min = min(n), Max = max(n)) |>
  mutate(Relationship = "Teachers per school")

students_per_school <- timss |>
  count(school_id, name = "n") |>
  summarise(Mean = mean(n), SD = sd(n), Min = min(n), Max = max(n)) |>
  mutate(Relationship = "Students per school")

bind_rows(students_per_teacher, teachers_per_school, students_per_school) |>
  dplyr::select(Relationship, Mean, SD, Min, Max) |>
  gt() |>
  tab_header(
    title    = "Cluster size variation",
    subtitle = "Data: TIMSS 2015 Grade 4 (Canada)"
  ) |>
  fmt_number(columns = c(Mean, SD), decimals = 1) |>
  fmt_number(columns = c(Min, Max), decimals = 0) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_title()
  )
```

## Outcome: math achievement

```{r}
#| message: false
#| warning: false
#| label: tbl-math-summary
#| tbl-cap: "Descriptive statistics for math achievement."

math_stats <- timss |>
  summarise(
    Mean = mean(math_score, na.rm = TRUE),
    SD   = sd(math_score,   na.rm = TRUE),
    Min  = min(math_score,  na.rm = TRUE),
    Max  = max(math_score,  na.rm = TRUE)
  )

math_stats |>
  gt() |>
  tab_header(
    title    = "Math achievement",
    subtitle = "Data: TIMSS 2015 Grade 4 (Canada)"
  ) |>
  fmt_number(columns = everything(), decimals = 1) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_title()
  )
```

```{r}
#| message: false
#| warning: false
#| label: fig-math-dist
#| fig-cap: "Distribution of math achievement."

timss |>
  ggplot(aes(x = math_score)) +
  geom_histogram(bins = 30, fill = "#009DD7", color = "white", alpha = 0.9) +  # AIR blue
  labs(
    x = "TIMSS math score",
    y = "Number of students"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## Intraclass correlations (ICCs)

### School-level ICC

```{r}
#| message: false
#| warning: false
#| code-line-numbers: true

school_icc_model <- lmer(
  math_score ~ 1 + (1 | school_id),                              # <1>
  data = timss
)

school_vc     <- as.data.frame(VarCorr(school_icc_model))        # <2>
school_bw_var <- school_vc$vcov[school_vc$grp == "school_id"]    # <3>
school_wi_var <- school_vc$vcov[school_vc$grp == "Residual"]
school_icc    <- school_bw_var / (school_bw_var + school_wi_var) # <4>

school_icc
```

1. This fits an empty (intercept-only) random-intercept model. `(1 | school_id)` lets the intercept vary by school, so the model partitions the outcome variance into between-school and within-school components.
2. `VarCorr()` returns the estimated variance components, and `as.data.frame()` makes them easy to subset.
3. The `grp` column labels each component: the between-school variance and the residual (within-school) variance.
4. The ICC is the between-school share of the total variance.

### Teacher-level ICC

```{r}
#| message: false
#| warning: false

teacher_icc_model <- lmer(
  math_score ~ 1 + (1 | teacher_id),
  data = timss
)

teacher_vc     <- as.data.frame(VarCorr(teacher_icc_model))
teacher_bw_var <- teacher_vc$vcov[teacher_vc$grp == "teacher_id"]
teacher_wi_var <- teacher_vc$vcov[teacher_vc$grp == "Residual"]
teacher_icc    <- teacher_bw_var / (teacher_bw_var + teacher_wi_var)

teacher_icc
```

### Three-level ICC decomposition

```{r}
#| message: false
#| warning: false
#| code-line-numbers: true

three_level_model <- lmer(
  math_score ~ 1 + (1 | school_id) + (1 | school_id:teacher_id), # <1>
  data = timss
)

three_vc          <- as.data.frame(VarCorr(three_level_model))
school_var        <- three_vc$vcov[three_vc$grp == "school_id"]
teacher_var       <- three_vc$vcov[three_vc$grp == "school_id:teacher_id"]
residual_var      <- three_vc$vcov[three_vc$grp == "Residual"]
total_var         <- school_var + teacher_var + residual_var

school_icc_3l    <- school_var   / total_var                     # <2>
teacher_icc_3l   <- teacher_var  / total_var

school_icc_3l
teacher_icc_3l
```

1. `(1 | school_id:teacher_id)` nests teachers within schools, so the model partitions the variance into school, teacher-within-school, and residual components.
2. Each level's ICC is that level's share of the total variance across the three components.

### Covariate ICCs

```{r}
#| message: false
#| warning: false

icc_for <- function(var) {
  fit <- lmer(reformulate("(1 | school_id)", response = var), data = timss)
  performance::icc(fit)$ICC_adjusted
}

covariate_iccs <- tibble(
  variable = c("math_score", "student_math_conf", "student_like_math",
               "student_age", "dad_edu", "mom_edu")
) |>
  mutate(icc = map_dbl(variable, icc_for))
```

```{r}
#| message: false
#| warning: false
#| label: tbl-covariate-iccs
#| tbl-cap: "School-level ICCs for the outcome and selected student covariates."

covariate_iccs |>
  gt() |>
  cols_label(variable = "Variable", icc = "School-level ICC") |>
  tab_header(
    title    = "Covariate ICCs",
    subtitle = "Data: TIMSS 2015 Grade 4 (Canada)"
  ) |>
  fmt_number(columns = icc, decimals = 3) |>
  tab_style(
    style     = cell_text(font = "monospace"),
    locations = cells_body(columns = variable)
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_title()
  )
```
